DaLi: Database as a Library

### 2017-02-19 paper/intro.tex
Intro expanded.

  \item In this paper we describe a library called \name that adresses
  the aforementioned problems. \name lets programmers obtain add
  persistence and fully-decentralized replication to their
  applications without ever having to leave the comforts of high-level
  reasoning in OCaml. Thus, in keeping with the long tradition of
  Functional Programming, programmers are \emph{completely} shielded
  from the reality of the underlying system implementation. This
  elevation from the low-level reasoning to the high-level reasoning
  concretely manifests as reasoning in terms of \emph{mergeable} data
  types instead of \emph{replicated} data types; replication being a
  system-level concern.

>>>>

  \item In this paper we describe an OCaml library called \name that
  adresses the aforementioned problems. \name lets programmers add
  persistence and fully-decentralized replication to their
  applications without ever having to leave the comforts of high-level
  reasoning in OCaml. Thus, in keeping with the long tradition of
  Functional Programming, programmers are \emph{completely} shielded
  from the complexities of the underlying system model. This elevation
  from the low-level reasoning to the high-level reasoning concretely
  manifests as reasoning in terms of \emph{mergeable} data types
  instead of \emph{replicated} data types; replication being a
  system-level concern. A mergeable data type library is a standard
  OCaml data type library equipped with a \C{merge} function that
  describes how to merge two concurrent versions of the data type that
  originated from a single common ancestor. Thus, any OCaml data type
  for which a meaningful merge function can be written is a mergeable
  type. We give multiple examples of mergeable types useful in
  practice, present a simple theory that guides programmers in writing
  merge functions, and describe the tool support available to them in
  this process. \name also comes with a novel programming model that
  realizes a monadic functional version control system centered around
  data, rather than file coherence. The \name monad lets programmers
  write and compose concurrent computations around multiple versions
  of a mergeable data type. The computation makes progress by forking
  concurrent versions of the existing versions, creating later
  versions of the existing versions along a \emph{branch}, or by
  merging existing versions across the branches. However, as we shall
  see, unconstrained branching may yield a branching structure that
  becomes unmergeable, thus leading to \emph{stuck} states. An
  important property of the \name library is that it suitably
  constrains the branching structure to avoid stuck states without
  introducing additional coordination that compromises the
  decentralized nature of the computation. Futhermore, \name
  profitably exploits the provenance information available via the
  branching structure to make the application resilient to network
  partitions. As a result, OCaml programmers get the benefits of
  persistence and highly available replication, while they continue to
  enjoy the comforts of high-level reasoning and the familiarity of
  the standard library data types.

  Our contributions are summarized below:
  \begin{itemize}
    \item We formally introduce the concept of a mergeable data type,
    as an alternative to a replicated data type, to admit high-level
    reasoning about replicated data in functional programs.

    \item We describe the \name library that transparently adds
    persistence and replication features to a mergeable type. \name
    comes with a programming model that brings to bear the power and
    flexibility of a version control system on administration of the
    replicated data. \name however hides the complexity of version
    control behind a monad, and exposes its functionality via a simple
    API that lets programmers define and compose concurrent OCaml
    computations around replicated data.

    \item We present a formalization of the \name programming model,
    and prove results that guarantee eventual convergence of
    concurrent versions, and progress (stuck-freeness) guarantees.

    \item We describe an implementation of \name atop the Irmin store,
    and describe multiple examples, and two case studies that
    attest to the practical utility of \name.
  \end{itemize}
